Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-12 write-ups/PLDI/paper.tex
Comments from Paul.

\newcommand{\offerfork}{\mathsf{offer\_fork}}
\newcommand{\offerkill}{\mathsf{offer\_kill}}
\newcommand{\globalbarrier}{\mathsf{global\_barrier}}
\newcommand{\resizingglobalbarrier}{\mathsf{resizing\_global\_barrier}}
\newcommand{\getgroupid}{\mathsf{get\_group\_id}}
\newcommand{\getnumgroups}{\mathsf{get\_num\_groups}}
\newcommand{\getlocalid}{\mathsf{get\_local\_id}}
\newcommand{\getglobalid}{\mathsf{get\_global\_id}}
\newcommand{\getlocalsize}{\mathsf{get\_global\_id}}
\newcommand{\getglobalsize}{\mathsf{get\_global\_size}}

@@

To operate correctly, a blocking concurrent algorithm requires
\emph{fair} scheduling of threads.  In the context of irregular
data-parallel algorithms accelerated on a GPU, this translates to
requiring fair scheduling of workgroups.  Without fairness, a
workgroup may fail to make progress, leading to starvation.  For
example, if workgroup $A$ holds a mutex, an unfair scheduler may cause
another workgroup $B$ to spin-wait forever for the mutex to be
released.  Similarly, an unfair scheduler can a workgroup to spin-wait
at a global barrier so that other workgroups do not make progress
towards the barrier.

@@

\item An experience report porting \NumAlgorithms{} irregular GPU
  algorithms to use cooperative kernels, and experimental results
  assessing the performance overhead associated with cooperation and
  the responsiveness to high priority tasks that cooperation enables,
  across four GPUs from two vendors (\mysec\ref{sec:experiments}).

@@

\paragraph{Memory spaces and memory model}
An OpenCL kernel has access to four memory spaces.  \emph{Shared virtual
memory} (SVM) is accessible to all threads and the host application concurrently.  \emph{Global} memory is shared among all threads executing a
kernel, but can only be safely accessed by the host application before
or after kernel execution.  Each workgroup has its own portion of \emph{local},
which can be used for fast intra-workgroup communication, and every thread has
an allocation of very fast \emph{private} memory, where
function-local variables are allocated.

Fine-grained communication between threads, and also with the host in
the case of SVM, is enabled via a set of atomic data types and
operations, based on the C11 memory model~\cite{C11}.

@@

HSA provides limited, one-way guarantees,
stating~\cite[p. 46]{HSAprogramming11}: \emph{Work-group A can wait
  for values written by work-group B without deadlock provided ... A
  comes after B in work-group flattened ID order}. However, this
one-way communication is not sufficient so support either the
work stealing or graph application examples of
Figures~\ref{fig:workstealing} and \ref{fig:graphsearch}. As described
above, both require \emph{symmetric} communication between
threads.

@@

(*@\label{line:graph:swap}@*)    swap(in_nodes, out_nodes);
(*@\label{line:graph:gb1}@*)    global_barrier();
(*@\label{line:graph:reset}@*)    reset(out_nodes);
    level++;
(*@\label{line:graph:gb2}@*)    global_barrier();
  }
}
\end{lstlisting}
\caption{An OpenCL graph traversal algorithm, using a global barrier for synchronisation}\label{fig:graphsearch}
\end{figure}

@@

Although not depicted here, concurrent accesses to each pool inside
$\mathsf{more\_work}$ and $\mathsf{pop\_or\_steal}$ are guarded by a
mutex, implemented using atomic compare and swap operations on
global memory.  As a result, the kernel presents two opportunities for
spin-waiting: spinning to obtain a mutex, and spinning in the main
kernel loop to obtain a task to process.  The algorithm thus depends
on a fair scheduler, which, as discussed in \mysec\ref{sec:intro},
is not provided by current GPU programming models.

@@

The application iterates as long as the current frontier contains
nodes to process (line~\ref{line:graph:iterate}). At each frontier,
the nodes to be processed are evenly distributed between the executing
threads. This is done through \emph{stride} based processing. A thread
retrieves its global thread id and the total number of threads
(obtained through the $\getglobalid()$ and $\getglobalsize()$ primitives). A thread then processes
each node associated with its id, i.e.\ nodes at indices that are a multiple of the total number of threads, plus the thread's id. A thread processes a node via the $\keyword{process\_node}$
function, which may (a) push nodes to process in the next frontier to
$\keyword{out\_nodes}$ and (b) use the current frontier, $\keyword{level}$, in
the computation. \ADComment{Tyler: should swap take the addresses of the pointers?} After processing the frontier, the threads swap their
local node array pointers (line~\ref{line:graph:swap}), i.e.\ the $\keyword{out\_nodes}$ of the current frontier becomes the $\keyword{in\_nodes}$ of the next
frontier.

@@

The global barrier used in this application is not provided as a GPU
primitive, though previous works have shown that such a global
barrier can be implemented~\cite{XF10,DBLP:conf/oopsla/SorensenDBGR16}. \ADComment{Are there additional papers we should be citing on GPU barriers?  E.g., is the sense reversal idea part of XF or is that from another paper?}  These implementations use well
defined atomic memory operations to ensure memory consistency, and
employ spinning to ensure threads wait at the barrier until all
threads have arrived.  This spinning behaviour requires fair
scheduling between workgroups for the barrier to operate correctly,
and as a result the barrier is \emph{not} guaranteed to work correctly
under current GPU programming models.  Such barriers appear to run
reliably on current GPUs for kernels that are executed with no more
workgroups than there are compute units; this is due to the fairness
of the occupancy-bound execution model that current GPUs appear to
provide.  But, as discussed in \mysec\ref{sec:intro}, this
occupancy-bound model is not mandated by language standards or vendor
implementations, and is unlikely to be guaranteed in the future.

>>>>

\newcommand{\offerfork}{\mathsf{offer\_fork}}
\newcommand{\offerkill}{\mathsf{offer\_kill}}
\newcommand{\globalbarrier}{\mathsf{global\_barrier}}
\newcommand{\resizingglobalbarrier}{\mathsf{resizing\_global\_barrier}}
\newcommand{\getgroupid}{\mathsf{get\_group\_id}}
\newcommand{\getnumgroups}{\mathsf{get\_num\_groups}}
\newcommand{\getlocalid}{\mathsf{get\_local\_id}}
\newcommand{\getglobalid}{\mathsf{get\_global\_id}}
\newcommand{\getlocalsize}{\mathsf{get\_local\_size}}
\newcommand{\getglobalsize}{\mathsf{get\_global\_size}}

@@

To operate correctly, a blocking concurrent algorithm requires
\emph{fair} scheduling of threads.  In the context of irregular
data-parallel algorithms accelerated on a GPU, this translates to
requiring fair scheduling of workgroups.  Without fairness, a
workgroup may fail to make progress, leading to starvation.  For
example, if workgroup $A$ holds a mutex, an unfair scheduler may cause
another workgroup $B$ to spin-wait forever for the mutex to be
released.  Similarly, an unfair scheduler can cause a workgroup to spin-wait
indefinitely at a global barrier so that other workgroups do not make progress
towards the barrier.

@@

\item An experience report of porting \NumAlgorithms{} irregular GPU
  algorithms to use cooperative kernels, and experimental results
  assessing the performance overhead associated with cooperation and
  the responsiveness to high priority tasks that cooperation enables,
  across four GPUs from two vendors (\mysec\ref{sec:experiments}).

@@

\paragraph{Memory spaces and memory model}
An OpenCL kernel has access to four memory spaces.  \emph{Shared virtual
memory} (SVM) is accessible to all threads and the host application concurrently.  \emph{Global} memory is shared among all threads executing a
kernel.  Each workgroup has its own portion of \emph{local} memory,
which can be used for fast intra-workgroup communication.  Every thread has
an allocation of very fast \emph{private} memory, where
function-local variables are allocated.

Communication between threads in the same workgroup can be achieved
using the workgroup barrier construct described above.  Finer-grained
communication within a workgroup, as well as inter-workgroup
communication and communication with the host while the kernel is
running, is enabled by a set of atomic data types and operations.  In
particular, fine-grained host/device communication is via atomic
operations on SVM.

@@

HSA provides limited, one-way guarantees,
stating~\cite[p. 46]{HSAprogramming11}: \emph{Work-group A can wait
  for values written by work-group B without deadlock provided ... A
  comes after B in work-group flattened ID order}. However, this
one-way communication is not sufficient to support blocking algorithms that use
mutexes and inter-workgroup barriers, both of which require \emph{symmetric} communication between
threads.

@@

(*@\label{line:graph:swap}@*)    swap(&in_nodes, &out_nodes);
(*@\label{line:graph:gb1}@*)    global_barrier();
(*@\label{line:graph:reset}@*)    reset(out_nodes);
    level++;
(*@\label{line:graph:gb2}@*)    global_barrier();
  }
}
\end{lstlisting}
\caption{An OpenCL graph traversal algorithm, using a global barrier for synchronisation}\label{fig:graphsearch}
\end{figure}

@@

Although not depicted here, concurrent accesses to each pool inside
$\mathsf{more\_work}$ and $\mathsf{pop\_or\_steal}$ are guarded by a
mutex, implemented using atomic compare and swap operations on
global memory.  As a result, the kernel presents two opportunities for
spin-waiting: spinning to obtain a mutex, and spinning in the main
kernel loop to obtain a task to process.  The algorithm thus depends
on a fair scheduler, which, as discussed in \mysec\ref{sec:intro},
is not guaranteed by current GPU programming models.

@@

The application iterates as long as the current frontier contains
nodes to process (line~\ref{line:graph:iterate}). At each frontier,
the nodes to be processed are evenly distributed between the executing
threads. This is done through \emph{stride} based processing. 
%
In this case, the stride is the total number of threads, obtained via 
$\getglobalsize()$.  Each thread processes nodes with the following sequence of indices:
\code{$\keyword{tid}$+$\keyword{stride}$*0},
\code{$\keyword{tid}$+$\keyword{stride}$*1},
\code{$\keyword{tid}$+$\keyword{stride}$*2}, etc., where $\keyword{tid}$ is a thread's global id.
A thread processes a node via the $\keyword{process\_node}$
function, which may (a) push nodes to process in the next frontier to
$\keyword{out\_nodes}$ and (b) use the current frontier, $\keyword{level}$, in
the computation. After processing the frontier, the threads swap their
local node array pointers (line~\ref{line:graph:swap}), i.e.\ the $\keyword{out\_nodes}$ of the current frontier becomes the $\keyword{in\_nodes}$ of the next
frontier.

@@

The global barrier used in this application is not provided as a GPU
primitive, though previous works have shown that such a global
barrier can be implemented~\cite{XF10,DBLP:conf/oopsla/SorensenDBGR16}. \ADComment{Are there additional papers we should be citing on GPU barriers?  E.g., is the sense reversal idea part of XF or is that from another paper?}  These implementations use well
defined atomic memory operations to ensure memory consistency, and
employ spinning to ensure threads wait at the barrier until all
threads have arrived.  This spinning behaviour requires fair
scheduling between workgroups for the barrier to operate correctly.
Again, fairness is not guaranteed by current GPU programming models.

The mutexes and barriers required by these two examples appear to run
reliably on current GPUs for kernels that are executed with no more
workgroups than there are compute units; this is due to the fairness
of the occupancy-bound execution model that current GPUs have been
shown, experimentally, to provide.  But, as discussed in
\mysec\ref{sec:intro}, this occupancy-bound model is not mandated by
language standards or vendor implementations, and is unlikely to be
guaranteed in the future.
